---
title: "Class Exercice 1"
output: html_notebook
---

Name:  
ID:  

# Diamonds price prediction

You are given two sets of data. The first is diamonds_data.csv which contains properties of several diamonds along with the selling price (Assume that this price is adjusted to present day value). Your objective is to predict a good price for each of the diamonds that are in the set diamonds_predictions.csv


1. Read the data that is in diamonds_dataset.csv and store it in DiamondsData

```{r}
DiamondsData<-read.csv("diamonds_dataset.csv", header = TRUE)
head(DiamondsData)
```

2. Develop a decision tree model to predict the price of diamonds


```{r}
library(rpart)
library(rpart.plot)

decTreeModel<-rpart(price~carat+cut+color+clarity+depth+table,data=DiamondsData)
prp(decTreeModel)
```




3. Read the data that is in diamonds_predictions.csv and store in decTreeDiamondsPredictions



```{r}
decTreeDiamondsPredictions<-read.csv("diamonds_predictions.csv", header = TRUE)
head(decTreeDiamondsPredictions)
```



4. Apply your decision tree model to predict the price of diamonds in the decTreeDiamondsPredictions data set. Store the predicted price in the price variable that is in the decTreeDiamondsPredictions.


```{r}
decTreeDiamondsPredictions$price <- predict(decTreeModel,decTreeDiamondsPredictions)
head(decTreeDiamondsPredictions)
```



5. Develop a regression model to predict the price diamonds

```{r}
#Removing cut, color, and clarity it does not offer numerical data and therefore, we cannot check if there is a correlation between it and the rest of the data. 
newDiamondsData = subset(DiamondsData, select= -c(cut,color,clarity))
head(newDiamondsData)
#Running the correlation table and removing features that are too correlated with each other. Removing price because we want to test for price. 
correlationTable = subset(newDiamondsData, select= -c(price))
head(correlationTable)
round(cor(correlationTable),2)
correlationTable = subset(correlationTable, select = -c(x))
round(cor(correlationTable),2)
correlationTable = subset(correlationTable, select = -c(y))
round(cor(correlationTable),2)
correlationTable = subset(correlationTable, select = -c(z))
round(cor(correlationTable),2)
#Correlation table now contains the features that are not correlated with each other
#Next is the creation of a regression model
regModel = lm(newDiamondsData$price ~ newDiamondsData$carat + newDiamondsData$depth + newDiamondsData$table)
summary(regModel)
#All variables are significant so all three variables will affect price in some way. 
```



6. Read the data that is in diamonds_predictions.csv and store in regressionDiamondsPredictions


```{r}
regressionDiamondsPredictions <- read.csv("diamonds_predictions.csv", header = TRUE)
head(regressionDiamondsPredictions)
```


7. Apply your regression model to predict the price of diamonds in the regressionDiamondsPredictions. data set. Store the predicted price in the price variable that is in the regressionDiamondsPredictions.

```{r}
nrow(regressionDiamondsPredictions)
regressionDiamondsPredictions$price <- predict(regModel,regressionDiamondsPredictions)
head(regressionDiamondsPredictions)
```



8. The actual prices are stored in diamonds_actual_prices.csv. Use these actual prices to evaluate the accuracy of the regression model and the decision tree model

```{r}
ActualPrices <- read.csv("diamonds_actual_prices.csv", header=TRUE)
head(ActualPrices)
decisionTreeError = decTreeDiamondsPredictions$price - ActualPrices
mseDecisionTree = mean(decisionTreeError^2)
print(mseDecisionTree)
regressionError = regressionDiamondsPredictions$price - ActualPrices
mseRegression = mean(regressionError^2)
print(mseRegression)

```




